---
title: "Chapter 7: Bonds"
subtitle: "Fixed-rate bond analytics with QuantLibGauss"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Fixed-rate bond analytics

This chapter builds a fixed-rate bond example using the current `QuantLibGauss` API.

The main policy in this chapter is:

- `qlg_fixed_rate_bond()` returns a `FixedRateBond` object directly.
- The bond schedule is built separately with `QuantLib::Schedule()`.
- `qlg_bond_price_measures()` receives `settlement_date_chr`, not a QuantLib Date object.
- `schedule = bond_schedule` is not passed to `qlg_bond_price_measures()`.
- `previous_coupon_date` and `accrued_days` are completed from the schedule afterward.
- Later TONA / z-spread / AssetSwap sections are kept as pending notes until the wrappers are fully aligned.

```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(QuantLib)
  library(QuantLibGauss)
})
```

# Local helpers

```{r local-helpers}
has_qlg <- function(fn) {
  fn %in% getNamespaceExports("QuantLibGauss")
}

show_tbl <- function(tbl, title = NULL, n = 20) {
  if (!is.null(title)) {
    cat("\n\n##", title, "\n\n")
  }

  tryCatch(
    {
      QuantLibGauss::qlg_show_tbl(tbl, title, n = n)
    },
    error = function(e) {
      print(utils::head(tbl, n))
    }
  )

  invisible(tbl)
}

metric_value_to_chr <- function(x) {
  if (length(x) == 0 || all(is.na(x))) {
    return(NA_character_)
  }

  x1 <- if (is.list(x)) x[[1]] else x[1]

  if (length(x1) == 0 || all(is.na(x1))) {
    return(NA_character_)
  }

  if (inherits(x1, "Date")) {
    return(as.character(x1))
  }

  if (is.numeric(x1)) {
    return(sprintf("%.8f", as.numeric(x1)))
  }

  as.character(x1)
}

metric_tbl_display <- function(tbl) {
  if (!"value" %in% names(tbl)) {
    return(tbl)
  }

  tbl |>
    dplyr::mutate(
      value = purrr::map_chr(value, metric_value_to_chr)
    )
}

get_metric_num <- function(tbl, metric_name) {
  if (!"metric" %in% names(tbl) || !"value" %in% names(tbl)) {
    return(NA_real_)
  }

  idx <- match(metric_name, tbl$metric)

  if (is.na(idx)) {
    return(NA_real_)
  }

  x <- tbl$value[[idx]]

  if (length(x) == 0 || all(is.na(x))) {
    return(NA_real_)
  }

  x1 <- if (is.list(x)) x[[1]] else x[1]

  suppressWarnings(as.numeric(x1))
}

replace_metric_value <- function(tbl, metric_name, new_value) {
  tbl <- tbl |>
    dplyr::mutate(value = as.list(value))

  idx <- match(metric_name, tbl$metric)

  if (is.na(idx)) {
    tbl <- dplyr::bind_rows(
      tbl,
      tibble::tibble(
        metric = metric_name,
        value = list(new_value)
      )
    )
  } else {
    tbl$value[idx] <- list(new_value)
  }

  tbl
}

make_thirty360 <- function() {
  attempts <- list(
    function() QuantLib::Thirty360("BondBasis"),
    function() QuantLib::Thirty360("EurobondBasis"),
    function() QuantLib::Thirty360("European"),
    function() QuantLib::Thirty360(0L),
    function() QuantLib::Thirty360(1L)
  )

  last_error <- NULL

  for (attempt in attempts) {
    out <- tryCatch(
      attempt(),
      error = function(e) {
        last_error <<- e
        NULL
      }
    )

    if (!is.null(out)) {
      return(out)
    }
  }

  stop(last_error)
}

complete_accrual_metrics <- function(
    price_tbl,
    bond_schedule,
    settlement_date_chr,
    settlement_date,
    day_counter
) {
  schedule_tbl <- QuantLibGauss::qlg_schedule_dates(bond_schedule)

  schedule_dates_chr <- schedule_tbl[[1]] |>
    as.character()

  schedule_dates_chr <- schedule_dates_chr[!is.na(schedule_dates_chr)]

  previous_coupon_date_chr <- max(
    schedule_dates_chr[schedule_dates_chr <= settlement_date_chr]
  )

  previous_coupon_date <- QuantLibGauss::qlg_date(previous_coupon_date_chr)

  accrued_days <- tryCatch(
    {
      day_counter$dayCount(previous_coupon_date, settlement_date)
    },
    error = function(e) {
      as.integer(round(
        day_counter$yearFraction(previous_coupon_date, settlement_date) * 360
      ))
    }
  )

  price_tbl |>
    replace_metric_value(
      metric_name = "previous_coupon_date",
      new_value = as.Date(previous_coupon_date_chr)
    ) |>
    replace_metric_value(
      metric_name = "accrued_days",
      new_value = as.numeric(accrued_days)
    )
}

bond_yield_from_clean_safe <- function(
    bond,
    clean_price,
    day_counter,
    compounding,
    frequency
) {
  attempts <- list(
    function() {
      QuantLibGauss::qlg_bond_yield_from_clean_price(
        bond = bond,
        clean_price = clean_price,
        day_counter = day_counter,
        compounding = compounding,
        frequency = frequency
      )
    },
    function() {
      QuantLibGauss::qlg_bond_yield(
        bond = bond,
        clean_price = clean_price,
        day_counter = day_counter,
        compounding = compounding,
        frequency = frequency
      )
    },
    function() {
      QuantLib::BondFunctions_yield(
        bond,
        clean_price,
        day_counter,
        compounding,
        frequency
      )
    }
  )

  last_error <- NULL

  for (attempt in attempts) {
    res <- tryCatch(
      attempt(),
      error = function(e) {
        last_error <<- e
        NULL
      }
    )

    if (!is.null(res)) {
      return(as.numeric(res))
    }
  }

  stop(last_error)
}
```

# Calendar, dates, and conventions

```{r enums-calendars-counters}
jp_calendar <- QuantLib::Japan()

dc_a365 <- QuantLib::Actual365Fixed()
dc_a360 <- QuantLib::Actual360()
dc_30 <- make_thirty360()

cmpd_compounded <- QuantLib::Compounding_Compounded_get()
cmpd_simple <- QuantLib::Compounding_Simple_get()

freq_annual <- QuantLib::Frequency_Annual_get()
freq_semiannual <- QuantLib::Frequency_Semiannual_get()

trade_date_chr <- "2022-08-19"
issue_date_chr <- "2022-07-28"
maturity_date_chr <- "2025-07-28"

trade_date <- QuantLibGauss::qlg_date(trade_date_chr)
issue_date_ql <- QuantLibGauss::qlg_date(issue_date_chr)
maturity_date_ql <- QuantLibGauss::qlg_date(maturity_date_chr)

settlement_days <- 2L

QuantLibGauss::qlg_set_eval_date(trade_date)

settlement_date <- QuantLibGauss::qlg_advance_days(
  calendar_obj = jp_calendar,
  date_obj = trade_date,
  n_days = settlement_days
)

settlement_date_chr <- QuantLibGauss::qlg_iso(settlement_date)

setup_tbl <- tibble::tibble(
  item = c(
    "trade_date",
    "settlement_date",
    "issue_date",
    "maturity_date",
    "settlement_days",
    "day_counter",
    "compounding",
    "frequency"
  ),
  value = c(
    trade_date_chr,
    settlement_date_chr,
    issue_date_chr,
    maturity_date_chr,
    as.character(settlement_days),
    "Thirty360(BondBasis)",
    "Compounded",
    "Semiannual"
  )
)

show_tbl(setup_tbl, "Bond setup dates and conventions")
```

# Fixed-rate bond setup

```{r fixed-rate-bond-setup}
face_amount <- 100
coupon_rate <- 0.00370
clean_price_input <- 97.0

bond_obj <- QuantLibGauss::qlg_fixed_rate_bond(
  issue_date = issue_date_chr,
  maturity_date = maturity_date_chr,
  face_amount = face_amount,
  coupon_rate = coupon_rate,
  settlement_days = settlement_days
)

bond_schedule <- QuantLib::Schedule(
  issue_date_ql,
  maturity_date_ql,
  QuantLibGauss::qlg_period_months(6),
  jp_calendar,
  "Unadjusted",
  "Unadjusted",
  "Backward",
  FALSE
)

schedule_tbl <- QuantLibGauss::qlg_schedule_dates(bond_schedule)

bond_setup_tbl <- tibble::tibble(
  metric = c(
    "bond_class",
    "face_amount",
    "coupon_rate",
    "clean_price_input"
  ),
  value = list(
    paste(class(bond_obj), collapse = ", "),
    face_amount,
    coupon_rate,
    clean_price_input
  )
)

show_tbl(metric_tbl_display(bond_setup_tbl), "Fixed-rate bond object")
show_tbl(schedule_tbl, "Bond schedule", n = 20)
```

# Clean price to yield

```{r clean-price-to-yield}
price_to_yield <- bond_yield_from_clean_safe(
  bond = bond_obj,
  clean_price = clean_price_input,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

yield_tbl <- tibble::tibble(
  metric = c(
    "clean_price_input",
    "yield_from_clean_price",
    "yield_percent"
  ),
  value = list(
    clean_price_input,
    price_to_yield,
    100 * price_to_yield
  )
)

show_tbl(metric_tbl_display(yield_tbl), "Clean price -> yield")
```

# Yield to price

```{r yield-to-price}
price_tbl <- QuantLibGauss::qlg_bond_price_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  settlement_date = settlement_date_chr
)

price_tbl <- complete_accrual_metrics(
  price_tbl = price_tbl,
  bond_schedule = bond_schedule,
  settlement_date_chr = settlement_date_chr,
  settlement_date = settlement_date,
  day_counter = dc_30
)

price_tbl_display <- metric_tbl_display(price_tbl)

show_tbl(price_tbl_display, "Yield -> price summary")

dirty_price_from_yield <- get_metric_num(price_tbl, "dirty_price")
clean_price_from_yield <- get_metric_num(price_tbl, "clean_price")
accrued_amount <- get_metric_num(price_tbl, "accrued_amount")

cat(
  "clean price:",
  sprintf("%.8f", clean_price_from_yield),
  "\n",
  "dirty price:",
  sprintf("%.8f", dirty_price_from_yield),
  "\n",
  "accrued:",
  sprintf("%.8f", accrued_amount),
  "\n"
)
```

# Risk measures

```{r risk-measures}
risk_tbl <- QuantLibGauss::qlg_bond_risk_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

risk_tbl_display <- metric_tbl_display(risk_tbl)

show_tbl(risk_tbl_display, "Bond risk measures")

modified_duration <- get_metric_num(risk_tbl, "modified_duration")
convexity_value <- get_metric_num(risk_tbl, "convexity")

cat(
  "modified duration:",
  sprintf("%.8f", modified_duration),
  "\n",
  "convexity:",
  sprintf("%.8f", convexity_value),
  "\n"
)
```

# Risk hand calculation

```{r risk-handcalc}
handcalc_tbl <- QuantLibGauss::qlg_bond_risk_handcalc(
  bond = bond_obj,
  yield = price_to_yield,
  dirty_price = dirty_price_from_yield,
  modified_duration = modified_duration,
  convexity = convexity_value,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

handcalc_tbl_display <- metric_tbl_display(handcalc_tbl)

show_tbl(handcalc_tbl_display, "Risk hand calculations")
```

# Manual cashflow present value check

```{r manual-cashflow-pv}
schedule_dates_chr <- schedule_tbl[[1]] |>
  as.character()

schedule_dates_chr <- schedule_dates_chr[!is.na(schedule_dates_chr)]

future_dates_chr <- schedule_dates_chr[schedule_dates_chr > settlement_date_chr]

coupon_per_period <- face_amount * coupon_rate / 2

manual_cf_tbl <- tibble::tibble(
  payment_date = future_dates_chr,
  period_no = seq_along(future_dates_chr),
  cashflow = coupon_per_period
) |>
  dplyr::mutate(
    cashflow = ifelse(
      .data$payment_date == maturity_date_chr,
      .data$cashflow + face_amount,
      .data$cashflow
    ),
    ql_date = purrr::map(payment_date, QuantLibGauss::qlg_date),
    time = purrr::map_dbl(
      ql_date,
      ~ dc_30$yearFraction(settlement_date, .x)
    ),
    discount_factor = 1 / (1 + price_to_yield / 2)^(2 * time),
    pv = cashflow * discount_factor
  ) |>
  dplyr::select(-ql_date)

manual_dirty_price <- sum(manual_cf_tbl$pv, na.rm = TRUE)
manual_clean_price <- manual_dirty_price - accrued_amount

manual_summary_tbl <- tibble::tibble(
  metric = c(
    "manual_dirty_price",
    "manual_clean_price",
    "wrapper_dirty_price",
    "wrapper_clean_price",
    "dirty_price_difference",
    "clean_price_difference"
  ),
  value = list(
    manual_dirty_price,
    manual_clean_price,
    dirty_price_from_yield,
    clean_price_from_yield,
    manual_dirty_price - dirty_price_from_yield,
    manual_clean_price - clean_price_from_yield
  )
)

show_tbl(manual_cf_tbl, "Manual cashflow PV table")
show_tbl(metric_tbl_display(manual_summary_tbl), "Manual PV check")
```

# Flat yield pricing engine

```{r flat-yield-pricing}
flat_yield_handle <- QuantLibGauss::qlg_bond_flat_forward_handle(
  settlement_date = settlement_date_chr,
  rate = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

flat_engine <- QuantLib::DiscountingBondEngine(flat_yield_handle)

QuantLib::Instrument_setPricingEngine(
  bond_obj,
  flat_engine
)

flat_dirty_price <- QuantLib::Instrument_NPV(bond_obj)

flat_yield_tbl <- tibble::tibble(
  metric = c(
    "flat_yield",
    "dirty_price_from_flat_engine",
    "dirty_price_from_yield",
    "difference"
  ),
  value = list(
    price_to_yield,
    flat_dirty_price,
    dirty_price_from_yield,
    flat_dirty_price - dirty_price_from_yield
  )
)

show_tbl(metric_tbl_display(flat_yield_tbl), "Flat yield pricing")
```

# Bond cashflow table

```{r bond-cashflow-table}
bond_cf_tbl <- tryCatch(
  {
    QuantLibGauss::qlg_bond_cashflow_table(
      bond = bond_obj,
      curve = flat_yield_handle
    )
  },
  error = function(e) {
    message("qlg_bond_cashflow_table(bond, curve) failed. Retrying with bond only.")

    QuantLibGauss::qlg_bond_cashflow_table(
      bond = bond_obj
    )
  }
)

show_tbl(bond_cf_tbl, "Bond cashflow table", n = 30)
```

# Bond analysis summary

```{r bond-analysis-summary}
summary_tbl <- tibble::tibble(
  metric = c(
    "trade_date",
    "settlement_date",
    "issue_date",
    "maturity_date",
    "clean_price_input",
    "yield_from_clean_price",
    "clean_price_from_yield",
    "dirty_price_from_yield",
    "accrued_amount",
    "modified_duration",
    "convexity"
  ),
  value = c(
    QuantLibGauss::qlg_iso(trade_date),
    settlement_date_chr,
    issue_date_chr,
    maturity_date_chr,
    sprintf("%.8f", clean_price_input),
    sprintf("%.8f", price_to_yield),
    sprintf("%.8f", clean_price_from_yield),
    sprintf("%.8f", dirty_price_from_yield),
    sprintf("%.8f", accrued_amount),
    sprintf("%.8f", modified_duration),
    sprintf("%.8f", convexity_value)
  )
)

show_tbl(summary_tbl, "Bond analysis summary", n = 20)

cat("\nchapter07 bonds stable rewrite completed successfully.\n")
```

# Later curve, spread, and asset-swap sections

```{r later-sections-pending}
available_later_functions <- tibble::tibble(
  function_name = getNamespaceExports("QuantLibGauss")
) |>
  dplyr::filter(
    stringr::str_detect(
      function_name,
      stringr::regex(
        "ois|tona|spread|zspread|z_spread|asset|swap|curve",
        ignore_case = TRUE
      )
    )
  ) |>
  dplyr::arrange(function_name)

show_tbl(
  available_later_functions,
  "Available QuantLibGauss functions for later sections",
  n = 100
)

later_sections_tbl <- tibble::tibble(
  section = c(
    "TONA curve",
    "Z-spread",
    "Asset swap"
  ),
  status = c(
    "pending",
    "pending",
    "deferred"
  ),
  note = c(
    "TONA / OIS curve section will be aligned after checking the current QuantLibGauss curve helper signatures.",
    "Z-spread wrappers exist, but this chapter first stabilizes fixed-rate bond analytics.",
    "AssetSwap is deferred until swap / index / curve wrappers are stable."
  )
)

show_tbl(
  later_sections_tbl,
  "Later curve / spread / asset-swap sections",
  n = 20
)
```

# Session information

```{r session-info}
sessionInfo()
```

