---
title: "Chapter 7: Bonds"
subtitle: "Fixed-rate bond analytics with QuantLibGauss"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Fixed-rate bond analytics

This chapter builds a fixed-rate bond example using the current `QuantLibGauss` API.

The main policy in this chapter is:

- `qlg_fixed_rate_bond()` returns a `FixedRateBond` object directly.
- The bond schedule is built separately with `QuantLib::Schedule()`.
- `qlg_bond_price_measures()` receives `settlement_date_chr`, not a QuantLib Date object.
- `schedule = bond_schedule` is not passed to `qlg_bond_price_measures()`.
- `previous_coupon_date` and `accrued_days` are completed from the schedule afterward.
- Later TONA / z-spread / AssetSwap sections are kept as pending notes until the wrappers are fully aligned.

```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(QuantLib)
  library(QuantLibGauss)
})
```

# Local helpers

```{r local-helpers}
has_qlg <- function(fn) {
  fn %in% getNamespaceExports("QuantLibGauss")
}

show_tbl <- function(tbl, title = NULL, n = 20) {
  if (!is.null(title)) {
    cat("\n\n##", title, "\n\n")
  }

  tryCatch(
    {
      QuantLibGauss::qlg_show_tbl(tbl, title, n = n)
    },
    error = function(e) {
      print(utils::head(tbl, n))
    }
  )

  invisible(tbl)
}

metric_value_to_chr <- function(x) {
  if (length(x) == 0 || all(is.na(x))) {
    return(NA_character_)
  }

  x1 <- if (is.list(x)) x[[1]] else x[1]

  if (length(x1) == 0 || all(is.na(x1))) {
    return(NA_character_)
  }

  if (inherits(x1, "Date")) {
    return(as.character(x1))
  }

  if (is.numeric(x1)) {
    return(sprintf("%.8f", as.numeric(x1)))
  }

  as.character(x1)
}

metric_tbl_display <- function(tbl) {
  if (!"value" %in% names(tbl)) {
    return(tbl)
  }

  tbl |>
    dplyr::mutate(
      value = purrr::map_chr(value, metric_value_to_chr)
    )
}

get_metric_num <- function(tbl, metric_name) {
  if (!"metric" %in% names(tbl) || !"value" %in% names(tbl)) {
    return(NA_real_)
  }

  idx <- match(metric_name, tbl$metric)

  if (is.na(idx)) {
    return(NA_real_)
  }

  x <- tbl$value[[idx]]

  if (length(x) == 0 || all(is.na(x))) {
    return(NA_real_)
  }

  x1 <- if (is.list(x)) x[[1]] else x[1]

  suppressWarnings(as.numeric(x1))
}

replace_metric_value <- function(tbl, metric_name, new_value) {
  tbl <- tbl |>
    dplyr::mutate(value = as.list(value))

  idx <- match(metric_name, tbl$metric)

  if (is.na(idx)) {
    tbl <- dplyr::bind_rows(
      tbl,
      tibble::tibble(
        metric = metric_name,
        value = list(new_value)
      )
    )
  } else {
    tbl$value[idx] <- list(new_value)
  }

  tbl
}

make_thirty360 <- function() {
  attempts <- list(
    function() QuantLib::Thirty360("BondBasis"),
    function() QuantLib::Thirty360("EurobondBasis"),
    function() QuantLib::Thirty360("European"),
    function() QuantLib::Thirty360(0L),
    function() QuantLib::Thirty360(1L)
  )

  last_error <- NULL

  for (attempt in attempts) {
    out <- tryCatch(
      attempt(),
      error = function(e) {
        last_error <<- e
        NULL
      }
    )

    if (!is.null(out)) {
      return(out)
    }
  }

  stop(last_error)
}

complete_accrual_metrics <- function(
    price_tbl,
    bond_schedule,
    settlement_date_chr,
    settlement_date,
    day_counter
) {
  schedule_tbl <- QuantLibGauss::qlg_schedule_dates(bond_schedule)

  schedule_dates_chr <- schedule_tbl[[1]] |>
    as.character()

  schedule_dates_chr <- schedule_dates_chr[!is.na(schedule_dates_chr)]

  previous_coupon_date_chr <- max(
    schedule_dates_chr[schedule_dates_chr <= settlement_date_chr]
  )

  previous_coupon_date <- QuantLibGauss::qlg_date(previous_coupon_date_chr)

  accrued_days <- tryCatch(
    {
      day_counter$dayCount(previous_coupon_date, settlement_date)
    },
    error = function(e) {
      as.integer(round(
        day_counter$yearFraction(previous_coupon_date, settlement_date) * 360
      ))
    }
  )

  price_tbl |>
    replace_metric_value(
      metric_name = "previous_coupon_date",
      new_value = as.Date(previous_coupon_date_chr)
    ) |>
    replace_metric_value(
      metric_name = "accrued_days",
      new_value = as.numeric(accrued_days)
    )
}

bond_yield_from_clean_safe <- function(
    bond,
    clean_price,
    day_counter,
    compounding,
    frequency
) {
  attempts <- list(
    function() {
      QuantLibGauss::qlg_bond_yield_from_clean_price(
        bond = bond,
        clean_price = clean_price,
        day_counter = day_counter,
        compounding = compounding,
        frequency = frequency
      )
    },
    function() {
      QuantLibGauss::qlg_bond_yield(
        bond = bond,
        clean_price = clean_price,
        day_counter = day_counter,
        compounding = compounding,
        frequency = frequency
      )
    },
    function() {
      QuantLib::BondFunctions_yield(
        bond,
        clean_price,
        day_counter,
        compounding,
        frequency
      )
    }
  )

  last_error <- NULL

  for (attempt in attempts) {
    res <- tryCatch(
      attempt(),
      error = function(e) {
        last_error <<- e
        NULL
      }
    )

    if (!is.null(res)) {
      return(as.numeric(res))
    }
  }

  stop(last_error)
}
```

# Calendar, dates, and conventions

```{r enums-calendars-counters}
jp_calendar <- QuantLib::Japan()

dc_a365 <- QuantLib::Actual365Fixed()
dc_a360 <- QuantLib::Actual360()
dc_30 <- make_thirty360()

cmpd_compounded <- QuantLib::Compounding_Compounded_get()
cmpd_simple <- QuantLib::Compounding_Simple_get()

freq_annual <- QuantLib::Frequency_Annual_get()
freq_semiannual <- QuantLib::Frequency_Semiannual_get()

trade_date_chr <- "2022-08-19"
issue_date_chr <- "2022-07-28"
maturity_date_chr <- "2025-07-28"

trade_date <- QuantLibGauss::qlg_date(trade_date_chr)
issue_date_ql <- QuantLibGauss::qlg_date(issue_date_chr)
maturity_date_ql <- QuantLibGauss::qlg_date(maturity_date_chr)

settlement_days <- 2L

QuantLibGauss::qlg_set_eval_date(trade_date)

settlement_date <- QuantLibGauss::qlg_advance_days(
  calendar_obj = jp_calendar,
  date_obj = trade_date,
  n_days = settlement_days
)

settlement_date_chr <- QuantLibGauss::qlg_iso(settlement_date)

setup_tbl <- tibble::tibble(
  item = c(
    "trade_date",
    "settlement_date",
    "issue_date",
    "maturity_date",
    "settlement_days",
    "day_counter",
    "compounding",
    "frequency"
  ),
  value = c(
    trade_date_chr,
    settlement_date_chr,
    issue_date_chr,
    maturity_date_chr,
    as.character(settlement_days),
    "Thirty360(BondBasis)",
    "Compounded",
    "Semiannual"
  )
)

show_tbl(setup_tbl, "Bond setup dates and conventions")
```

# Fixed-rate bond setup

```{r fixed-rate-bond-setup}
face_amount <- 100
coupon_rate <- 0.00370
clean_price_input <- 97.0

bond_obj <- QuantLibGauss::qlg_fixed_rate_bond(
  issue_date = issue_date_chr,
  maturity_date = maturity_date_chr,
  face_amount = face_amount,
  coupon_rate = coupon_rate,
  settlement_days = settlement_days
)

bond_schedule <- QuantLib::Schedule(
  issue_date_ql,
  maturity_date_ql,
  QuantLibGauss::qlg_period_months(6),
  jp_calendar,
  "Unadjusted",
  "Unadjusted",
  "Backward",
  FALSE
)

schedule_tbl <- QuantLibGauss::qlg_schedule_dates(bond_schedule)

bond_setup_tbl <- tibble::tibble(
  metric = c(
    "bond_class",
    "face_amount",
    "coupon_rate",
    "clean_price_input"
  ),
  value = list(
    paste(class(bond_obj), collapse = ", "),
    face_amount,
    coupon_rate,
    clean_price_input
  )
)

show_tbl(metric_tbl_display(bond_setup_tbl), "Fixed-rate bond object")
show_tbl(schedule_tbl, "Bond schedule", n = 20)
```

# Clean price to yield

```{r clean-price-to-yield}
price_to_yield <- bond_yield_from_clean_safe(
  bond = bond_obj,
  clean_price = clean_price_input,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

yield_tbl <- tibble::tibble(
  metric = c(
    "clean_price_input",
    "yield_from_clean_price",
    "yield_percent"
  ),
  value = list(
    clean_price_input,
    price_to_yield,
    100 * price_to_yield
  )
)

show_tbl(metric_tbl_display(yield_tbl), "Clean price -> yield")
```

# Yield to price

```{r yield-to-price}
price_tbl <- QuantLibGauss::qlg_bond_price_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  settlement_date = settlement_date_chr
)

price_tbl <- complete_accrual_metrics(
  price_tbl = price_tbl,
  bond_schedule = bond_schedule,
  settlement_date_chr = settlement_date_chr,
  settlement_date = settlement_date,
  day_counter = dc_30
)

price_tbl_display <- metric_tbl_display(price_tbl)

show_tbl(price_tbl_display, "Yield -> price summary")

dirty_price_from_yield <- get_metric_num(price_tbl, "dirty_price")
clean_price_from_yield <- get_metric_num(price_tbl, "clean_price")
accrued_amount <- get_metric_num(price_tbl, "accrued_amount")

cat(
  "clean price:",
  sprintf("%.8f", clean_price_from_yield),
  "\n",
  "dirty price:",
  sprintf("%.8f", dirty_price_from_yield),
  "\n",
  "accrued:",
  sprintf("%.8f", accrued_amount),
  "\n"
)
```

# Risk measures

```{r risk-measures}
risk_tbl <- QuantLibGauss::qlg_bond_risk_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

risk_tbl_display <- metric_tbl_display(risk_tbl)

show_tbl(risk_tbl_display, "Bond risk measures")

modified_duration <- get_metric_num(risk_tbl, "modified_duration")
convexity_value <- get_metric_num(risk_tbl, "convexity")

cat(
  "modified duration:",
  sprintf("%.8f", modified_duration),
  "\n",
  "convexity:",
  sprintf("%.8f", convexity_value),
  "\n"
)
```

# Risk hand calculation

```{r risk-handcalc}
handcalc_tbl <- QuantLibGauss::qlg_bond_risk_handcalc(
  bond = bond_obj,
  yield = price_to_yield,
  dirty_price = dirty_price_from_yield,
  modified_duration = modified_duration,
  convexity = convexity_value,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

handcalc_tbl_display <- metric_tbl_display(handcalc_tbl)

show_tbl(handcalc_tbl_display, "Risk hand calculations")
```

# Manual cashflow present value check

```{r manual-cashflow-pv}
schedule_dates_chr <- schedule_tbl[[1]] |>
  as.character()

schedule_dates_chr <- schedule_dates_chr[!is.na(schedule_dates_chr)]

future_dates_chr <- schedule_dates_chr[schedule_dates_chr > settlement_date_chr]

coupon_per_period <- face_amount * coupon_rate / 2

manual_cf_tbl <- tibble::tibble(
  payment_date = future_dates_chr,
  period_no = seq_along(future_dates_chr),
  cashflow = coupon_per_period
) |>
  dplyr::mutate(
    cashflow = ifelse(
      .data$payment_date == maturity_date_chr,
      .data$cashflow + face_amount,
      .data$cashflow
    ),
    ql_date = purrr::map(payment_date, QuantLibGauss::qlg_date),
    time = purrr::map_dbl(
      ql_date,
      ~ dc_30$yearFraction(settlement_date, .x)
    ),
    discount_factor = 1 / (1 + price_to_yield / 2)^(2 * time),
    pv = cashflow * discount_factor
  ) |>
  dplyr::select(-ql_date)

manual_dirty_price <- sum(manual_cf_tbl$pv, na.rm = TRUE)
manual_clean_price <- manual_dirty_price - accrued_amount

manual_summary_tbl <- tibble::tibble(
  metric = c(
    "manual_dirty_price",
    "manual_clean_price",
    "wrapper_dirty_price",
    "wrapper_clean_price",
    "dirty_price_difference",
    "clean_price_difference"
  ),
  value = list(
    manual_dirty_price,
    manual_clean_price,
    dirty_price_from_yield,
    clean_price_from_yield,
    manual_dirty_price - dirty_price_from_yield,
    manual_clean_price - clean_price_from_yield
  )
)

show_tbl(manual_cf_tbl, "Manual cashflow PV table")
show_tbl(metric_tbl_display(manual_summary_tbl), "Manual PV check")
```

# Flat yield pricing engine

```{r flat-yield-pricing}
flat_yield_handle <- QuantLibGauss::qlg_bond_flat_forward_handle(
  settlement_date = settlement_date_chr,
  rate = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

flat_engine <- QuantLib::DiscountingBondEngine(flat_yield_handle)

QuantLib::Instrument_setPricingEngine(
  bond_obj,
  flat_engine
)

flat_dirty_price <- QuantLib::Instrument_NPV(bond_obj)

flat_yield_tbl <- tibble::tibble(
  metric = c(
    "flat_yield",
    "dirty_price_from_flat_engine",
    "dirty_price_from_yield",
    "difference"
  ),
  value = list(
    price_to_yield,
    flat_dirty_price,
    dirty_price_from_yield,
    flat_dirty_price - dirty_price_from_yield
  )
)

show_tbl(metric_tbl_display(flat_yield_tbl), "Flat yield pricing")
```

# Bond cashflow table

```{r bond-cashflow-table}
bond_cf_tbl <- tryCatch(
  {
    QuantLibGauss::qlg_bond_cashflow_table(
      bond = bond_obj,
      curve = flat_yield_handle
    )
  },
  error = function(e) {
    message("qlg_bond_cashflow_table(bond, curve) failed. Retrying with bond only.")

    QuantLibGauss::qlg_bond_cashflow_table(
      bond = bond_obj
    )
  }
)

show_tbl(bond_cf_tbl, "Bond cashflow table", n = 30)
```

# Bond analysis summary

```{r bond-analysis-summary}
summary_tbl <- tibble::tibble(
  metric = c(
    "trade_date",
    "settlement_date",
    "issue_date",
    "maturity_date",
    "clean_price_input",
    "yield_from_clean_price",
    "clean_price_from_yield",
    "dirty_price_from_yield",
    "accrued_amount",
    "modified_duration",
    "convexity"
  ),
  value = c(
    QuantLibGauss::qlg_iso(trade_date),
    settlement_date_chr,
    issue_date_chr,
    maturity_date_chr,
    sprintf("%.8f", clean_price_input),
    sprintf("%.8f", price_to_yield),
    sprintf("%.8f", clean_price_from_yield),
    sprintf("%.8f", dirty_price_from_yield),
    sprintf("%.8f", accrued_amount),
    sprintf("%.8f", modified_duration),
    sprintf("%.8f", convexity_value)
  )
)

show_tbl(summary_tbl, "Bond analysis summary", n = 20)

cat("\nchapter07 bonds stable rewrite completed successfully.\n")
```

# Later curve, spread, and asset-swap sections

```{r later-sections-pending}
available_later_functions <- tibble::tibble(
  function_name = getNamespaceExports("QuantLibGauss")
) |>
  dplyr::filter(
    stringr::str_detect(
      function_name,
      stringr::regex(
        "ois|tona|spread|zspread|z_spread|asset|swap|curve",
        ignore_case = TRUE
      )
    )
  ) |>
  dplyr::arrange(function_name)

show_tbl(
  available_later_functions,
  "Available QuantLibGauss functions for later sections",
  n = 100
)

later_sections_tbl <- tibble::tibble(
  section = c(
    "TONA curve",
    "Z-spread",
    "Asset swap"
  ),
  status = c(
    "pending",
    "pending",
    "deferred"
  ),
  note = c(
    "TONA / OIS curve section will be aligned after checking the current QuantLibGauss curve helper signatures.",
    "Z-spread wrappers exist, but this chapter first stabilizes fixed-rate bond analytics.",
    "AssetSwap is deferred until swap / index / curve wrappers are stable."
  )
)

show_tbl(
  later_sections_tbl,
  "Later curve / spread / asset-swap sections",
  n = 20
)
```

# Session information

```{r session-info}
sessionInfo()
```
## このノートの目的

この章では、Python / QuantLib notebook の Bonds 章を、`QuantiveRiemann` と `QuantLib(SWIG for R)` を前提に R で書き直す。

扱う内容は次の通りである。

1. fixed-rate bond setup  
2. clean price から yield への変換  
3. yield から clean / dirty price への変換  
4. duration / BPV / convexity  
5. hand calculation checks  
6. bond cashflow table  
7. flat-yield pricing  
8. TONA curve pricing  
9. z-spread  
10. asset swap  

## 注意

- `bondYield()` は QuantLib 1.39 以降で clean price wrapper が必要な build がある。
- `zSpread()` も同様に wrapper が必要な場合がある。
- `AssetSwap()` は RFR index 使用時、floating schedule を明示指定する方が安全である。

## 前提

- `# devtools::load_all(".")  # disabled for Quarto book` 済み、または package install 済み
- `utility.R`, `wrapper_lib.R`, `bond.R` の関数が使える
- QuantLib SWIG build で Bond / OIS / AssetSwap まわりが利用可能

```{r setup}
# ============================================================
# 0. setup
# ============================================================


suppressPackageStartupMessages({
  library(tidyverse)
})

```

## 1. Enums / calendars / counters

```{r enums-calendars-counters}
jp_calendar <- Japan()

dc_a365 <- Actual365Fixed()
dc_a360 <- Actual360()
dc_30 <- Thirty360("BondBasis")

cmpd_compounded <- Compounding_Compounded_get()
cmpd_simple <- Compounding_Simple_get()

freq_annual <- Frequency_Annual_get()
freq_semiannual <- Frequency_Semiannual_get()
```

## 2. Fixed-rate bond setup

ここでは、固定利付債を作成する。  
評価日は 2022-08-19、約定から 2 営業日後を settlement date とする。

```{r fixed-rate-bond-setup}
trade_date <- QuantLibGauss::qlg_date("2022-08-19")
settlement_days <- 2L

effective_date <- QuantLibGauss::qlg_date("2022-07-28")
maturity_date <- QuantLibGauss::qlg_date("2025-07-28")

face_amount <- 100
coupon_rate <- 0.00370
clean_price_input <- 97.0

QuantLibGauss::qlg_set_eval_date(trade_date)

settlement_date <- QuantLibGauss::qlg_advance_days(
  calendar_obj = jp_calendar,
  date_obj = trade_date,
  n_days = settlement_days
)

bond_res <- QuantLibGauss::qlg_fixed_rate_bond(
  trade_date = trade_date,
  settlement_days = settlement_days,
  effective_date = effective_date,
  maturity_date = maturity_date,
  face_amount = face_amount,
  coupon_rate = coupon_rate,
  calendar = jp_calendar,
  schedule_frequency = QuantLibGauss::qlg_period_months(6),
  accrual_day_counter = dc_30,
  payment_convention = "Unadjusted",
  maturity_convention = "Unadjusted",
  date_generation = "Backward",
  end_of_month = FALSE
)

bond_schedule <- bond_res$schedule
bond_obj <- bond_res$bond

QuantLibGauss::qlg_show_tbl(
  QuantLibGauss::qlg_schedule_dates(bond_schedule),
  "Bond schedule",
  n = 20
)
```

## 3. Clean price から yield

clean price 97.0 から、債券利回りを逆算する。

```{r clean-price-to-yield}
price_to_yield <- QuantLibGauss::qlg_bond_yield_from_clean_price(
  bond = bond_obj,
  clean_price = clean_price_input,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

cat(
  "Price 97.0 -> yield:",
  sprintf("%.6f%%", 100 * price_to_yield),
  "\n"
)
```

## 4. Yield から clean / dirty price

求めた yield を使って、clean price と dirty price を再計算する。

```{r yield-to-price}
price_tbl <- QuantLibGauss::qlg_bond_price_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  settlement_date = settlement_date,
  schedule = bond_schedule
)

QuantLibGauss::qlg_show_tbl(
  price_tbl,
  "Yield -> price summary",
  n = 20
)

get_metric_value <- function(tbl, metric_name) {
  tbl |>
    dplyr::filter(metric == metric_name) |>
    dplyr::pull(value) |>
    purrr::pluck(1) |>
    as.numeric()
}

dirty_price_from_yield <- get_metric_value(price_tbl, "dirty_price")
clean_price_from_yield <- get_metric_value(price_tbl, "clean_price")
accrued_amount <- get_metric_value(price_tbl, "accrued_amount")
```

## 5. Risk measures

Duration、BPV、Convexity などのリスク指標を確認する。

```{r risk-measures}
risk_tbl <- QuantLibGauss::qlg_bond_risk_measures(
  bond = bond_obj,
  yield = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

QuantLibGauss::qlg_show_tbl(
  risk_tbl,
  "Bond risk measures",
  n = 20
)

modified_duration <- risk_tbl |>
  dplyr::filter(metric == "modified_duration") |>
  dplyr::pull(value) |>
  as.numeric()

convexity_value <- risk_tbl |>
  dplyr::filter(metric == "convexity") |>
  dplyr::pull(value) |>
  as.numeric()
```

## 6. Hand calculation checks

QuantLib の計算結果に対して、手計算ベースの確認を行う。

```{r hand-calculation-checks}
handcalc_tbl <- QuantLibGauss::qlg_bond_risk_handcalc(
  bond = bond_obj,
  yield = price_to_yield,
  dirty_price = dirty_price_from_yield,
  modified_duration = modified_duration,
  convexity = convexity_value,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

QuantLibGauss::qlg_show_tbl(
  handcalc_tbl,
  "Hand calculations",
  n = 20
)
```

## 7. Bond cashflow table under flat yield

flat yield curve を作り、そのカーブで債券キャッシュフローを割り引く。

```{r cashflow-table-flat-yield}
flat_yield_handle <- QuantLibGauss::qlg_bond_flat_forward_handle(
  settlement_date = settlement_date,
  rate = price_to_yield,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

flat_engine <- DiscountingBondEngine(flat_yield_handle)
Instrument_setPricingEngine(bond_obj, flat_engine)

bond_cf_tbl_flat <- QuantLibGauss::qlg_bond_cashflow_table(
  bond = bond_obj,
  curve = flat_yield_handle
)

QuantLibGauss::qlg_show_tbl(
  bond_cf_tbl_flat,
  "Bond cashflow table (flat yield discounting)",
  n = 20
)

cat(
  "Hand dirty price from CF table:",
  sprintf("%.8f", sum(bond_cf_tbl_flat$pv, na.rm = TRUE)),
  "\n"
)
```

### Macaulay duration の手計算

```{r macaulay-duration-hand}
macaulay_hand <- bond_cf_tbl_flat |>
  dplyr::filter(!is.na(pv), pay_date > QuantLibGauss::qlg_iso(settlement_date)) |>
  dplyr::mutate(
    year_fraction = purrr::map_dbl(
      pay_date,
      ~ dc_30$yearFraction(settlement_date, QuantLibGauss::qlg_date(.x))
    )
  ) |>
  dplyr::summarise(value = sum(year_fraction * pv) / dirty_price_from_yield) |>
  dplyr::pull(value)

cat(
  "Macaulay duration (hand):",
  sprintf("%.8f", macaulay_hand),
  "\n"
)

bond_cf_2025_tbl <- bond_cf_tbl_flat |>
  dplyr::filter(as.Date(pay_date) >= as.Date("2025-01-01"))

QuantLibGauss::qlg_show_tbl(
  bond_cf_2025_tbl,
  "Bond cashflows from 2025 onward",
  n = 20
)
```

## 8. TONA curve for spread analysis

ここでは spread analysis 用に TONA curve を作成する。  
パッケージ側に curve builder が入った場合は、この関数は差し替えてよい。

```{r tona-curve-builder}
make_tona_curve <- function(curve_data, trade_date = "2022-08-19") {
  trade_date_ql <- QuantLibGauss::qlg_date(trade_date)
  QuantLibGauss::qlg_set_eval_date(trade_date_ql)

  tona_curve_handle <- RelinkableYieldTermStructureHandle()

  tona_index <- OvernightIndex(
    "TONA",
    0,
    JPYCurrency(),
    jp_calendar,
    dc_a365,
    tona_curve_handle
  )

  helper_list <- purrr::map(curve_data, function(x) {
    kind <- as.character(x[[1]])
    tenor <- as.character(x[[2]])
    rate_pct <- as.numeric(x[[3]])

    if (kind == "depo") {
      return(
        DepositRateHelper(
          QuoteHandle(SimpleQuote(rate_pct / 100)),
          tona_index
        )
      )
    }

    if (kind == "swap") {
      return(
        OISRateHelper(
          2,
          QuantLibGauss::qlg_period(tenor),
          QuoteHandle(SimpleQuote(rate_pct / 100)),
          tona_index
        )
      )
    }

    stop("Unsupported helper type: ", kind)
  })

  helper_vec <- RateHelperVector()
  purrr::walk(helper_list, ~ RateHelperVector_append(helper_vec, .x))

  tona_curve_obj <- PiecewiseLogLinearDiscount(
    0,
    jp_calendar,
    helper_vec,
    dc_a365
  )

  TermStructure_enableExtrapolation(tona_curve_obj)
  RelinkableYieldTermStructureHandle_linkTo(tona_curve_handle, tona_curve_obj)

  list(
    index = tona_index,
    curve = tona_curve_obj,
    curve_handle = tona_curve_handle,
    quote_tbl = tibble::tibble(
      kind = purrr::map_chr(curve_data, 1),
      tenor = purrr::map_chr(curve_data, 2),
      rate_pct = purrr::map_dbl(curve_data, ~ as.numeric(.x[[3]]))
    )
  )
}
```

```{r build-tona-curve}
tona_curve_data <- list(
  c("depo", "1d", -0.00900),
  c("swap", "1m", -0.01807),
  c("swap", "6m", -0.01043),
  c("swap", "12m", 0.01250),
  c("swap", "18m", 0.03125),
  c("swap", "2y", 0.04875),
  c("swap", "3y", 0.07375),
  c("swap", "5y", 0.11854),
  c("swap", "7y", 0.19146)
)

tona_bundle <- make_tona_curve(
  tona_curve_data,
  trade_date = "2022-08-19"
)

QuantLibGauss::qlg_show_tbl(tona_bundle$quote_tbl, "TONA input quotes")
QuantLibGauss::qlg_show_tbl(QuantLibGauss::qlg_curve_tbl(tona_bundle$curve), "TONA curve table")
```

### I-spread の簡易確認

満期時点に対応する TONA zero rate を補間し、bond yield との差を見る。

```{r i-spread-check}
curve_proxy_tbl <- QuantLibGauss::qlg_curve_tbl(tona_bundle$curve, n = 500)

mat_time <- dc_a365$yearFraction(
  tona_bundle$curve$referenceDate(),
  maturity_date
)

interp_rate <- approx(
  curve_proxy_tbl$time,
  curve_proxy_tbl$zero,
  xout = mat_time,
  rule = 2
)$y

cat(
  "Interpolated TONA rate at maturity:",
  sprintf("%.6f%%", 100 * interp_rate),
  ", bond yield:",
  sprintf("%.6f%%", 100 * price_to_yield),
  ", I-spread:",
  sprintf("%.6f%%", 100 * (price_to_yield - interp_rate)),
  "\n"
)
```

## 9. Discounting with flat forward and TONA curve

同じ債券を、flat forward と TONA curve でそれぞれ評価する。

```{r discounting-flat-tona}
flat_forward_price <- QuantLibGauss::qlg_bond_npv_with_flat_yield(
  bond = bond_obj,
  settlement_date = settlement_date,
  rate = 0.01418713,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

cat(
  "Dirty price with 1.418713% flat yield:",
  sprintf("%.6f", flat_forward_price),
  "\n"
)

tona_discount_price <- QuantLibGauss::qlg_bond_npv_with_curve(
  bond = bond_obj,
  curve_handle = tona_bundle$curve_handle
)

cat(
  "Dirty price discounted on TONA curve:",
  sprintf("%.6f", tona_discount_price),
  "\n"
)
```

## 10. Z-spread / zero-spreaded curve

TONA curve に対して、clean price 97.0 を再現する z-spread を求める。

```{r z-spread}
z_spread_value <- QuantLibGauss::qlg_bond_zspread(
  bond = bond_obj,
  clean_price = 97.0,
  curve = tona_bundle$curve,
  day_counter = dc_30,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

cat(
  "z-spread for clean price 97.0:",
  sprintf("%.6f%%", 100 * z_spread_value),
  "\n"
)

z_spread_price_res <- QuantLibGauss::qlg_bond_npv_with_zspread(
  bond = bond_obj,
  base_curve_handle = tona_bundle$curve_handle,
  z_spread = z_spread_value,
  compounding = cmpd_compounded,
  frequency = freq_annual,
  day_counter = dc_a365
)

cat(
  "Dirty price on TONA + z-spread curve:",
  sprintf("%.6f", z_spread_price_res$npv),
  "\n"
)

bond_cf_tbl_spread <- QuantLibGauss::qlg_bond_cashflow_table(
  bond = bond_obj,
  curve = z_spread_price_res$spread_curve_handle
)

QuantLibGauss::qlg_show_tbl(
  bond_cf_tbl_spread,
  "Bond cashflow table (TONA + z-spread)",
  n = 20
)

cat(
  "Hand price on TONA + z-spread:",
  sprintf("%.6f", sum(bond_cf_tbl_spread$pv, na.rm = TRUE)),
  "\n"
)
```

## 11. AS spread scenarios

Asset swap spread の理解のために、簡易的に TONA discount price と annuity から AS spread を計算する。

```{r as-spread-scenarios}
bond_cf_tbl_tona <- QuantLibGauss::qlg_bond_cashflow_table(
  bond = bond_obj,
  curve = tona_bundle$curve_handle
) |>
  dplyr::filter(cashflow_type == "coupon")

yr_frac_vec <- purrr::map_dbl(
  bond_cf_tbl_tona$pay_date,
  ~ dc_a365$yearFraction(settlement_date, QuantLibGauss::qlg_date(.x))
)

tenor_flow <- diff(c(0, yr_frac_vec))
tona_annuity <- sum(bond_cf_tbl_tona$discount_factor * tenor_flow, na.rm = TRUE)

benchmark_dirty_price <- 90.0256

as_spread_input_tbl <- tibble::tibble(
  price_label = c("model_dirty_price", "benchmark_dirty_price"),
  target_dirty_price = c(dirty_price_from_yield, benchmark_dirty_price)
)

as_spread_result_tbl <- as_spread_input_tbl |>
  dplyr::mutate(
    discount_price = tona_discount_price,
    annuity = tona_annuity,
    as_spread = (discount_price - target_dirty_price) / annuity
  )

QuantLibGauss::qlg_show_tbl(
  tibble::tibble(
    metric = c("tona_discount_price", "tona_annuity"),
    value = c(tona_discount_price, tona_annuity)
  ),
  "AS spread base inputs",
  n = 20
)

QuantLibGauss::qlg_show_tbl(
  as_spread_result_tbl,
  "AS spread scenarios",
  n = 20
)
```

## 12. AssetSwap

最後に `AssetSwap()` を用いた分析を行う。  
元コードでは `clean_price_asw`, `credit_spread_asw`, `pay_fixed_rate`, `is_par_asset_swap` が未定義だったため、ここでは実行可能なようにデフォルト値を明示する。

```{r asset-swap-default-inputs}
clean_price_asw <- clean_price_input
credit_spread_asw <- 0.0
pay_fixed_rate <- TRUE
is_par_asset_swap <- TRUE

tibor6m_for_asw <- Tibor(
  QuantLibGauss::qlg_period_months(6),
  tona_bundle$curve_handle
)
```

```{r asset-swap-analysis}
asset_swap_tbl <- QuantLibGauss::qlg_asset_swap_analysis(
  bond = bond_obj,
  clean_price = clean_price_asw,
  ibor_index = tibor6m_for_asw,
  spread = credit_spread_asw,
  settlement_date = settlement_date,
  maturity_date = maturity_date,
  calendar = jp_calendar,
  floating_schedule_frequency = QuantLibGauss::qlg_period_years(1),
  payment_convention = "Unadjusted",
  date_generation = "Backward",
  end_of_month = FALSE,
  floating_day_counter = dc_a365,
  pay_fixed_rate = pay_fixed_rate,
  par_asset_swap = is_par_asset_swap,
  discount_curve_handle = tona_bundle$curve_handle
)

asset_swap_tbl_display <- asset_swap_tbl |>
  dplyr::mutate(
    value = purrr::map2_chr(metric, value, function(metric, x) {
      if (length(x) == 0 || all(is.na(x))) {
        return(NA_character_)
      }

      if (metric == "fair_spread") {
        return(sprintf("%.6f%%", 100 * as.numeric(x[[1]])))
      }

      if (metric == "fair_clean_price") {
        return(sprintf("%.6f", as.numeric(x[[1]])))
      }

      as.character(x[[1]])
    })
  )

QuantLibGauss::qlg_show_tbl(
  asset_swap_tbl_display,
  "AssetSwap result",
  n = 20
)

cat("\nchapter04 bonds rewrite completed successfully.\n")
```

## 実行方法

この `.qmd` は、プロジェクト直下で次のように実行できる。

```bash
quarto render chapter04_bonds_rewrite.qmd
```

HTML だけを出したい場合は、YAML の `format:` から `pdf:` を削除してもよい。
