---
title: "Chapter 8: Treasury Futures and CTD"
subtitle: "Deliverable basket, basis, implied repo, and CTD analysis with GaussQuant"
format:
  html:
    toc: true
    number-sections: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

# Treasury futures and cheapest-to-deliver analysis

This chapter studies a US Treasury futures deliverable basket with the current `GaussQuant` API.

The workflow is:

- build a Treasury bond and verify its price and accrued interest;
- calculate gross basis, carry, net basis, and implied repo;
- select the cheapest-to-deliver bond from the minimum net basis;
- calculate the CTD-adjusted futures BPV.

```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(QuantLib)
  library(GaussQuant)
})
```

# Calendar, dates, and conventions

```{r treasury-futures-setup}
cal_us_gov <- QuantLib::UnitedStates("GovernmentBond")

dc_act_act_bond <- GaussQuant::day_counter_GQL("ActualActual_Bond")
dc_act_360 <- GaussQuant::day_counter_GQL("Actual360")

cmpd_compounded <- QuantLib::Compounding_Compounded_get()
freq_semiannual <- QuantLib::Frequency_Semiannual_get()

settlement_days_t1 <- 1L
trade_date_chr <- "2023-04-20"
repo_end_date_chr <- "2023-07-06"

trade_date <- GaussQuant::date_GQL(trade_date_chr)
repo_end_date_ql <- GaussQuant::date_GQL(repo_end_date_chr)

GaussQuant::set_eval_date_GQL(trade_date)

settle_date <- GaussQuant::advance_days_GQL(
  calendar_obj = cal_us_gov,
  date_obj = trade_date,
  n_days = settlement_days_t1
)

settle_date_chr <- as.character(settle_date)
settle_date_ql <- GaussQuant::date_GQL(settle_date_chr)

setup_tbl <- tibble::tribble(
  ~item, ~value,
  "trade_date", trade_date_chr,
  "spot_settlement_date", settle_date_chr,
  "repo_end_date", repo_end_date_chr,
  "settlement_days", as.character(settlement_days_t1),
  "calendar", "UnitedStates(GovernmentBond)",
  "bond_day_counter", "ActualActual(Bond)",
  "repo_day_counter", "Actual360",
  "compounding", "Compounded",
  "frequency", "Semiannual"
)

GaussQuant::show_tbl_GQH(setup_tbl, "Treasury futures setup")
```

# US Treasury bond example

```{r us-treasury-bond}
bond_bundle <- GaussQuant::us_treasury_bond_GQL(
  effective_date = "2022-09-30",
  maturity_date = "2027-09-30",
  coupon_rate_pct = 4.125,
  face_amount = 100,
  settlement_days = settlement_days_t1,
  calendar = cal_us_gov,
  day_counter = dc_act_act_bond
)

bond_obj <- bond_bundle$bond

GaussQuant::show_tbl_GQH(
  GaussQuant::schedule_dates_GQL(bond_bundle$schedule),
  "US Treasury bond schedule",
  n = 20
)

clean_price_at_six_percent <- GaussQuant::bond_clean_price_from_yield_safe_GQL(
  bond = bond_obj,
  yield_rate = 6 / 100,
  day_counter = dc_act_act_bond,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  settlement_date = GaussQuant::date_GQL("2023-07-06")
)

cat(
  "Yield 6%, settle 2023-07-06 clean price:",
  GaussQuant::fmt_num_GQH(clean_price_at_six_percent, 4),
  "\n"
)
```

# Bond price from yield

```{r bond-price-from-yield}
bond_yield <- 3.70 / 100
futures_price <- 109 + 10 / 32

accrued_amount <- bond_obj$accruedAmount(settle_date_ql)

clean_price <- GaussQuant::bond_clean_price_from_yield_safe_GQL(
  bond = bond_obj,
  yield_rate = bond_yield,
  day_counter = dc_act_act_bond,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  settlement_date = settle_date_ql
)

dirty_price <- clean_price + accrued_amount

price_tbl <- tibble::tibble(
  trade_date = as.Date(trade_date_chr),
  settlement_date = as.Date(settle_date_chr),
  yield = bond_yield,
  accrued_amount = accrued_amount,
  clean_price = clean_price,
  dirty_price = dirty_price,
  futures_price = futures_price
)

GaussQuant::show_tbl_GQH(price_tbl, "Bond price from yield")

cat(
  "Yield:",
  sprintf("%.8f", bond_yield),
  "\nClean price:",
  sprintf("%.8f", clean_price),
  "\nDirty price:",
  sprintf("%.8f", dirty_price),
  "\nAccrued amount:",
  sprintf("%.8f", accrued_amount),
  "\n"
)
```

# Bond cashflow present-value illustration

The generic interest-rate discount factors below are illustrative. They do not exactly reproduce QuantLib bond-yield pricing because ActualActual(Bond) uses coupon reference periods.

```{r bond-cashflow-pv}
interest_rate_obj <- QuantLib::InterestRate(
  bond_yield,
  dc_act_act_bond,
  cmpd_compounded,
  freq_semiannual
)

settle_date_r <- as.Date(settle_date_chr)

bond_cf_tbl <- GaussQuant::bond_cashflow_table_GQL(bond_obj) |>
  dplyr::mutate(
    pay_date = as.Date(.data$date),
    discount_factor = purrr::map_dbl(
      as.character(.data$pay_date),
      function(x) {
        if (as.Date(x) < settle_date_r) {
          return(NA_real_)
        }

        tryCatch(
          interest_rate_obj$discountFactor(
            settle_date_ql,
            GaussQuant::date_GQL(x)
          ),
          error = function(e) NA_real_
        )
      }
    ),
    pv = .data$amount * .data$discount_factor
  ) |>
  dplyr::select(
    .data$pay_date,
    .data$amount,
    .data$discount_factor,
    .data$pv
  )

cashflow_pv <- sum(bond_cf_tbl$pv, na.rm = TRUE)

cashflow_check_tbl <- tibble::tribble(
  ~metric, ~value,
  "cashflow_present_value", cashflow_pv,
  "dirty_price_from_yield", dirty_price,
  "illustrative_difference", cashflow_pv - dirty_price
)

GaussQuant::show_tbl_GQH(bond_cf_tbl, "Bond cashflow present-value table", n = 20)
GaussQuant::show_tbl_GQH(
  GaussQuant::metric_tbl_display_GQH(cashflow_check_tbl),
  "Cashflow PV illustration"
)
```

# Accrued interest hand calculation

```{r accrued-interest-handcalc}
previous_coupon_date <- GaussQuant::date_GQL("2023-03-30")
next_coupon_date <- GaussQuant::date_GQL("2023-09-30")

accrued_days_hand <- dc_act_act_bond$dayCount(
  previous_coupon_date,
  settle_date_ql
)

coupon_period_days <- dc_act_act_bond$dayCount(
  previous_coupon_date,
  next_coupon_date
)

coupon_amount_semiannual <- 4.125 / 2

accrued_interest_hand <- coupon_amount_semiannual *
  accrued_days_hand /
  coupon_period_days

accrued_tbl <- tibble::tibble(
  previous_coupon_date = as.Date(GaussQuant::iso_GQL(previous_coupon_date)),
  settlement_date = as.Date(settle_date_chr),
  next_coupon_date = as.Date(GaussQuant::iso_GQL(next_coupon_date)),
  accrued_days = accrued_days_hand,
  coupon_period_days = coupon_period_days,
  coupon_amount = coupon_amount_semiannual,
  accrued_interest_hand = accrued_interest_hand,
  accrued_amount_quantlib = accrued_amount,
  difference = accrued_interest_hand - accrued_amount
)

GaussQuant::show_tbl_GQH(accrued_tbl, "Accrued interest hand calculation")
```

# Deliverable basket

```{r deliverable-basket}
deliverable_tbl <- tibble::tribble(
  ~issue_date, ~maturity_date, ~coupon_rate_pct, ~conversion_factor, ~market_yield_pct,
  "2022-09-30", "2027-09-30", 4.125, 0.9305, 3.70,
  "2022-08-31", "2027-08-31", 3.125, 0.8953, 3.69,
  "2023-01-31", "2028-01-31", 3.500, 0.9011, 3.65
)

GaussQuant::show_tbl_GQH(deliverable_tbl, "Treasury futures deliverable basket")
```

# Gross basis, net basis, and CTD

```{r treasury-futures-ctd}
repo_rate <- 5.10 / 100

repo_days <- dc_act_360$dayCount(
  settle_date_ql,
  repo_end_date_ql
)

repo_year_fraction <- dc_act_360$yearFraction(
  settle_date_ql,
  repo_end_date_ql
)

ctd_result <- GaussQuant::bond_futures_ctd_table_GQL(
  deliverable_tbl = deliverable_tbl,
  settlement_date = settle_date_chr,
  futures_price = futures_price,
  repo_end_date = repo_end_date_chr,
  repo_rate = repo_rate,
  settlement_days = settlement_days_t1,
  calendar = cal_us_gov,
  day_counter = dc_act_act_bond,
  compounding = cmpd_compounded,
  frequency = freq_semiannual,
  repo_day_counter = dc_act_360,
  carry_day_counter = dc_act_360
)

basket_tbl <- ctd_result$basket

gross_basis_tbl <- basket_tbl |>
  dplyr::select(
    .data$issue_date,
    .data$maturity,
    .data$coupon,
    .data$yield_pct,
    .data$bpv,
    .data$clean_price,
    .data$dirty_price,
    .data$conversion_factor,
    .data$gross_basis
  )

net_basis_tbl <- basket_tbl |>
  dplyr::select(
    .data$maturity,
    .data$conversion_factor,
    .data$gross_basis,
    .data$accrued_start,
    .data$accrued_end,
    .data$coupon_income,
    .data$repo_cost,
    .data$carry,
    .data$net_basis,
    .data$forward_price,
    .data$implied_repo,
    .data$ctd
  )

basket_display_tbl <- basket_tbl |>
  dplyr::select(-dplyr::any_of("bond_obj"))

ctd_tbl <- ctd_result$ctd
ctd_display_tbl <- ctd_tbl |>
  dplyr::select(-dplyr::any_of("bond_obj"))

stopifnot(
  nrow(ctd_tbl) == 1L,
  isTRUE(ctd_tbl$ctd[[1L]])
)

cat(
  "Futures price:",
  sprintf("%.6f", futures_price),
  "\nRepo rate:",
  sprintf("%.6f", repo_rate),
  "\nRepo days:",
  repo_days,
  "\nRepo year fraction:",
  sprintf("%.8f", repo_year_fraction),
  "\n"
)

GaussQuant::show_tbl_GQH(gross_basis_tbl, "Gross basis table", n = 20)
GaussQuant::show_tbl_GQH(net_basis_tbl, "Net basis and implied repo table", n = 20)
GaussQuant::show_tbl_GQH(basket_display_tbl, "Deliverable basket summary", n = 20)
GaussQuant::show_tbl_GQH(ctd_display_tbl, "Cheapest-to-deliver bond", n = 20)
```

# CTD-adjusted futures BPV

```{r ctd-futures-bpv}
ctd_bond <- ctd_tbl$bond_obj[[1L]]
ctd_conversion_factor <- ctd_tbl$conversion_factor[[1L]]
ctd_yield <- ctd_tbl$yield_pct[[1L]] / 100

ctd_futures_bpv <- GaussQuant::bond_futures_bpv_GQL(
  bond = ctd_bond,
  conversion_factor = ctd_conversion_factor,
  contracts = 1,
  measure = "dv01",
  ytm = ctd_yield,
  day_counter = dc_act_act_bond,
  compounding = cmpd_compounded,
  frequency = freq_semiannual
)

ctd_summary_tbl <- tibble::tibble(
  metric = c(
    "ctd_maturity",
    "ctd_coupon",
    "ctd_yield_pct",
    "conversion_factor",
    "gross_basis",
    "net_basis",
    "implied_repo",
    "ctd_bond_bpv",
    "ctd_futures_bpv"
  ),
  value = list(
    as.character(ctd_tbl$maturity[[1L]]),
    ctd_tbl$coupon[[1L]],
    ctd_tbl$yield_pct[[1L]],
    ctd_conversion_factor,
    ctd_tbl$gross_basis[[1L]],
    ctd_tbl$net_basis[[1L]],
    ctd_tbl$implied_repo[[1L]],
    ctd_tbl$bpv[[1L]],
    ctd_futures_bpv
  )
)

GaussQuant::show_tbl_GQH(
  GaussQuant::metric_tbl_display_GQH(ctd_summary_tbl),
  "CTD and futures BPV summary"
)

cat(
  "\nChapter 08 treasury futures and CTD migrated to GaussQuant successfully.\n"
)
```
